Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-06 analysis/paper/paper.Rmd
edit discussion section

```{r PCs-distribution, fig.cap= "The distribution of mormalized PC scores by phases with CV values (%)"}

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and the Chinese periods, and between the pre-European and Post-European periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. This can be also supported by an increase in body diameter of vessels. Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape. Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods. That means pottery shape became more homogeneous and standardized after contact with the Chinese. A test of morphological disparity we conducted as a cross validation that also supports this result. In addition, we found a larger and more standardized size, indicated by body diameter of vessels, in the Chinese period. We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period using body diameter as a proxy variable of size. This might hint the intent to achieve a homogeneous form of vessels, but we still need to obtain more metric variables to assess the variation in size. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization. We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape. However, the effect can be reduced using CV, which offers a way to robustly compare variations across uneven sample sizes, regardless of a smaller sample size in the whole dataset [@Eerkens2001, pp.498]. Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well.

>>>>

```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases with CV values (%)"}

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans with a high diversity of types [@Cheng2008; @Wang2011]. This hinted at the emergence of social inequality within the Indigenous community. To understand the relationship between social inequality and foreign presences, we examined ceramic vessel shape standardization to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and the pre-European and Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Such shape is more pronounced in the Chinese period. In general, vessels become oval-shaped with a restricted mouth over time, which corresponds with an increase in body diameter of vessels. The correlation between shape and body diameter suggests that body diameter significantly varies with vessel shape. Oval-shaped vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods. A test of morphological disparity also demonstrates a similar result as well (See Online Supplementary Materials), suggesting that pottery shape became more standardized after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even at small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization [@Costin2001; @Costin1995]. In other words, mechanical standardization depends on potter's skills and experiences that increased uniformity normally reflects fewer producers.

To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
